Exploring how changes in cause of death have narrowed the black-white life expectancy gap in the United States.

### 2017-02-15 Manuscript/Manuscript_COD_decomp.Rmd
update std mort plots

###Introduction
The life expectancy of a population is a barometer for overall health. In 2014, life expectancy in the United States was 78.7 years, but hidden within this figure are differences by race. Non-Hispanic whites lived 3.6 years longer than Non-Hispanic blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf]. While both black and white Americans have experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied. During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, due to differential improvements in heart disease [Harper 2007]. The gap continued to narrow during the 2000s, but states differed markedly in the magnitude of improvement. For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [Harper 2014]. This study seeks to better understand these state-level differences, by examining how changes in the cause of death over time have worked to narrow or exacerbate the life expectancy gap within each state between 1969 and 2013. Common reasons for the decreasing life expectancy gap can thereby be identified, and anomalous states can be examined. Ideally, one would like to identify policies or social conditions in highly successful states that help to explain their unusual performance.

@@

```{r prep_era_figure_males, echo = F}
year1 <- 1969
year2 <- 1983
year3 <- 1993
year4 <- 2013

@@

>>>>

###Introduction
The life expectancy of a population is a barometer for overall health. In 2014, life expectancy in the United States was 78.8 years, but hidden within this figure are differences by race. Whites lived 3.4 years longer than Blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf]. While both black and white Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied. During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due largely to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, due to differential improvements in heart disease [Harper 2007]. The gap continued to narrow during the 2000s, but states differed markedly in the magnitude of improvement. For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [Harper 2014]. This study seeks to better understand these state-level differences, by examining how changes in the cause of death over time have worked to narrow or exacerbate the life expectancy gap within each state between 1969 and 2013. Common reasons for the decreasing life expectancy gap can thereby be identified, and anomalous states can be examined. Ideally, one would like to identify policies or social conditions in highly successful states that help to explain their unusual performance.

@@

```{r prep_era_figure, echo = F}
year1 <- 1969
year2 <- 1983
year3 <- 1993
year4 <- 2013

@@

For females, the life expectancy gap narrowed in `r dim(bw_dat.females[bw_dat.females$era1.diff > 0, ])[1]` of `r dim(bw_dat.females)[1]` states between 1969 and 1983. In `r length(cod.df$narrowed_gap[cod.df$sex == "Female" & cod.df$era == "1969-1983" & cod.df$COD == "Cardiovascular" & cod.df$narrowed_gap == T])` states, CVD made a substantial contribution towards narrowing the gap (for example, Virginia, Kentucky, South Carolina), while in others, its contribution was towards widening the gap (CI may be very important here -- the states with very large contributions also have very few blacks...).
